Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-01-21 doc/invt.tex
add a template configuration file

Another special case is the one-step process\cite{vankampen},
in each step, the system hops to the previous $i - 1$ bin
or the next bin $i + 1$ with equal probability $1/2$,
ignoring out-of-boundary transitions.
%
We shall further assume that
the eigenvectors of the $\mathbf w$ matrix
coincide with the those of the transition matrix
of the above sampling process
(cf. Sec. \ref{sec:band-matrix}).
%
Since the transition matrix is tridiagonal,
as given by Eq. \eqref{eq:wtridiag} with $\mu_1 = 1/2$,
the eigenvalues are given
by Eq. \eqref{eq:wband_eigenvalue} as
$$
\Lambda_i = 1 - 2 \sin^2 \frac{ (i - 1) \, \pi } { 2 \, n }.
$$
Note, however, these $\Lambda_i$'s are the eigenvalues of
the transition matrix of sampling scheme,
not to be confused with the eigenvalues, $\lambda_i$'s,
of the $\mathbf w$ matrix.
%
The correlation function $\kappa_i(t)$ of
decays as $\kappa_i(0) \, \Lambda_i^t$,
with the same-time value $\kappa_i(0) = 1$ being the same as
that in perfect sampling.
%
Thus, from Eq. \eqref{eq:Gamma_sum}, we get
%
\begin{equation}
\Gamma_i
=
1 + 2 \, \sum_{t = 1}^\infty \Lambda_i^t
=
\frac{ 1 + \Lambda_i } { 1 - \Lambda_i }
=
\cot^2 \left( \frac{ i \, \pi } { 2 \, n } \right).
\label{eq:Gamma_onestep}
\end{equation}
%
We can use this case to model local sampling processes.

@@

We now show
that the single-bin scheme
is an optimal scheme.
%
Consider a class of $\mathbf w$ matrices
sharing the same set of eigenvectors,
hence same $\Gamma_i$'s,
but with different eigenvalues, $\lambda_i$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of $\beta_i \ge 0$
%
\begin{align}
&
\left(
  \int_0^t dt'
    \sum_{i = 2}^n
      \Gamma_i \, \dot u_i^2\bigl( q(t') \bigr)
\right)
%
\left(
  \int_0^t dt'
    \sum_{i = 2}^n
      \Gamma_i \, c_i^2
\right)
%
\notag
\\
&
\ge
\left(
  \int_0^t dt'
    \sum_{i = 2}^n
      \Gamma_i \, c_i \, \dot u_i\bigl( q(t') \bigr)
\right)^2
\notag
\\
&
=
\left(
  \sum_{i = 2}^n \Gamma_k \, c_i
    \left[
      1 - e^{ -\lambda_i q(t) }
    \right]
\right)^2.
\notag
\end{align}
%
\note{A short proof of the above inequality
  is the following.
  %
  First, we have
$$
\int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \,
    \left( \dot u_i \, X - c_i \right)^2
\ge 0,
$$
  for any $X$.
  %
  Thus, the left-hand side,
  expanded as a quadratic polynomial of $X$,
  $A \, X^2 + B \, X + C$,
  must have a non-positive discriminant,
  $B^2 - 4 \, A \, C$.
  %
  This condition gives the desired inequality.
  \\%
}
%
Note that the last expression depends on
$q(t')$ only through the end point value,
which is fixed in the variation process.
%
Thus, the inequality sets a lower bound
of the error $\mathcal E$
expressed in Eq. \eqref{eq:error_mbin1},
%
The equality is achieved
if $\dot u_i\left( q(t') \right) = c_i$
for all $i = 2, \dots, n$ at any $t'$,
up to a multiplicative factor.
%
Solving this set of equations yields
$$
\alpha(t') = \frac{              1              }
                  { \lambda_i \, (t' + t_{i0} ) },
\quad
\mathrm{with\;\;}
t_{i0} = \frac{             t            }
              { e^{ \lambda_i q(t) } - 1 }.
$$
Such a solution is possible only if
\begin{equation}
\lambda_2 = \dots = \lambda_n.
\end{equation}
%
\note{
  We add some steps for solution.
  Integrating $\dot u_i = c_i$ yields
  \begin{equation}
  u_i(t') = c_i \left(t' + t_{i0} \right).
  \label{eq:ui_solution}
  \end{equation}
  Given that $u_i(t) = 1$ and $u_i(0) = e^{-\lambda_i \, q(t)}$,
  we have
  $$
  c_i = \frac{ 1 }{ t + t_{i0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{i0} } { t + t_{i0} }
  =
  e^{ -\lambda_k \, q(t) }.
  $$
  Taking the logarithm of Eq. \eqref{eq:ui_solution} yields
  $$
  -\lambda_i \, q(t) + \lambda_i \, q(t')
  = \log c_i + \log\left( t' + t_{i0} \right).
  $$
  Differentiating this with respect to $t'$ yields
  $$
  \lambda_i \, \alpha(t') = \frac{ 1 } { t' + t_{i0} }.
  $$
}

@@

If we wish to find,
for a given updating matrix $\mathbf w$,
a near-optimal $\alpha(t)$,
%
then a good candidate is
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_trial}
\end{equation}
%
and to find the optimal value of $\lambda$.
%
Using Eq. \eqref{eq:alpha_trial}
in Eq. \eqref{eq:error_mbin1} yields
%
\begin{align}
\mathcal E
&=
\frac{1}{t+t_0}
\sum_{i=2}^n
  \frac{       \Gamma_i \, \lambda_i^2      }
       { \lambda \, (2 \lambda_i - \lambda) }
\left[
  1 - \left(
        1 + \frac{ t }{ t_0 }
      \right)^{1 - 2 \, \lambda_i/\lambda}
\right]
\notag
\\
&\stackrel{      t \gg t_0      }
          { =\joinrel=\joinrel= }
%
\sum_{i = 2}^n
\begin{dcases}
  \frac{ 1 }
       { t }
  \frac{         \Gamma_i \, \lambda_i^2       }
       { \lambda \, (2 \, \lambda_i - \lambda) }
  &
  \mathrm{if \;} \lambda < 2 \, \lambda_i,
  \\%[1em]
  %
  %
  \frac{ \Gamma_i }
       {  4 \, t  }
  \ln \frac{ t + t_0 }
           {  t_0    }
  &
  \mathrm{if \;} \lambda = 2 \, \lambda_i,
  \\%[1em]
  %
  %
  \frac{ t_0^{ 2 \, \lambda_i / \lambda  - 1 } }
       {   t^{ 2 \, \lambda_i / \lambda      } }
  \frac{      \Gamma_i \, \lambda_i^2          }
       { \lambda \, (\lambda - 2 \, \lambda_i) }
  &
  \mathrm{if \;} \lambda > 2 \, \lambda_i.
\end{dcases}
\label{eq:y2_sum_perfect_invt}
\end{align}
%
\note{In the above expression,
$t + t_0$ are replaced by $t$.
%
Adding back $t_0$, we get
$$
\begin{aligned}
  \mathcal E
  \stackrel{         t \to \infty         }
           { =\joinrel=\joinrel=\joinrel= }
  \sum_{i = 2}^n
  \begin{dcases}
    \frac{    1    }
         { t + t_0 }
    \frac{       \Gamma_i \, \lambda_i^2         }
         { \lambda \, (2 \, \lambda_i - \lambda) }
    &
    \mathrm{if \;} \lambda < 2 \, \lambda_i,
    \\%[1em]
    %
    %
    \frac{    \Gamma_i    }
         { 4 \, (t + t_0) }
    \ln \frac{ t + t_0 }
             {   t_0  }
    &
    \mathrm{if \;} \lambda = 2 \, \lambda_i,
    \\
    %
    %
    \frac{  t_0^{ 2 \, \lambda_i / \lambda  - 1}  }
         { (t + t_0)^{ 2 \, \lambda_i / \lambda } }
    \frac{      \Gamma_i \, \lambda_i^2          }
         { \lambda \, (\lambda - 2 \, \lambda_i) }
    &
    \mathrm{if \;} \lambda > 2 \, \lambda_i.
  \end{dcases}
\end{aligned}
$$
}
%
The last two cases, $\lambda \ge 2\lambda_i$,
are asymptotically suboptimal
as they decay slower than $1/t$.
%
Thus, the optimal $\lambda$
satisfies $\lambda < 2 \min \lambda_i$,
and in this case,
we can seek that
the optimal value from
the smallest real root of
%
\begin{equation}
\sum_{i = 2}^n
\frac{ \lambda_i - \lambda }
{ \left(2 - \lambda/ \lambda_i \right)^2 }
= 0.
\label{eq:optimal_lambda_approx}
\end{equation}

>>>>

Another special case is the one-step process\cite{vankampen}:
in each step, the system hops to the previous bin $i - 1$
or the next bin $i + 1$ with equal probability $\frac 1 2$,
ignoring out-of-boundary transitions.
%
The transition matrix is
%
\begin{equation}
\arraycolsep=3.6pt\def\arraystretch{1.4}
\mathbf T
=
\left(
  \begin{array}{cccccccc}
    \frac 1 2 & \frac 1 2 & 0 & \dots & 0 \\
    \frac 1 2 & 0         & \frac 1 2 & \dots & 0 \\
    \vdots & &  & & \vdots \\
    0 & \dots & \frac 1 2 & 0  & \frac 1 2 \\
    0 & \dots & 0 & \frac 1 2 & \frac 1 2
  \end{array}
\right).
\label{eq:T_tridiag}
\end{equation}
%
We shall further assume that
the eigenvectors of $\mathbf w$
coincide with those of $\mathbf T$,
while the eigenvalues may differ
(this is true for a class of matrices
discussed in Sec. \ref{sec:band-matrix}).
%
The eigenvalues of $\mathbf T$ are
$$
\Lambda_i = 1 - 2 \sin^2 \frac{ (i - 1) \, \pi } { 2 \, n }.
$$
Note that these $\Lambda_i$'s are
not to be confused with the eigenvalues of $\mathbf w$,
$\lambda_i$'s.
%
The correlation function $\kappa_i(t)$
decays as $\kappa_i(0) \, \Lambda_i^t$,
with the same-time value $\kappa_i(0) = 1$ being the same as
that in perfect sampling.
%
Thus, from Eq. \eqref{eq:Gamma_sum}, we get
%
\begin{equation}
\Gamma_i
=
1 + 2 \, \sum_{t = 1}^\infty \Lambda_i^t
=
\frac{ 1 + \Lambda_i } { 1 - \Lambda_i }
=
\cot^2 \left[ \frac{ (i - 1) \, \pi } { 2 \, n } \right].
\label{eq:Gamma_onestep}
\end{equation}
%
We can use this case to model local sampling processes.

@@

We now show
that the single-bin scheme
is an optimal scheme.
%
Consider a class of $\mathbf w$ matrices
sharing the same set of eigenvectors,
hence same $\Gamma_i$'s,
but with different eigenvalues, $\lambda_i$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of $\beta_i \ge 0$
%
\begin{align}
&
\left(
  \int_0^t dt'
    \sum_{i = 2}^n
      \Gamma_i \, \dot u_i^2\bigl( q(t') \bigr)
\right)
%
\left(
  \int_0^t dt'
    \sum_{i = 2}^n
      \Gamma_i \, c_i^2
\right)
%
\notag
\\
&
\ge
\left(
  \int_0^t dt'
    \sum_{i = 2}^n
      \Gamma_i \, c_i \, \dot u_i\bigl( q(t') \bigr)
\right)^2
\notag
\\
&
=
\left(
  \sum_{i = 2}^n \Gamma_k \, c_i
    \left[
      1 - e^{ -\lambda_i q(t) }
    \right]
\right)^2.
\notag
\end{align}
%
\note{A short proof of the above inequality
  is the following.
  %
  First, we have
$$
\int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \,
    \left( \dot u_i \, X - c_i \right)^2
\ge 0,
$$
  for any $X$.
  %
  Thus, the left-hand side,
  expanded as a quadratic polynomial of $X$,
  $A \, X^2 + B \, X + C$,
  must have a non-positive discriminant,
  $B^2 - 4 \, A \, C$.
  %
  This condition gives the desired inequality.
  \\%
}
%
Note that the last expression depends on
$q(t')$ only through the end point value,
which is fixed in the variation process.
%
Thus, the inequality sets a lower bound
of the error $\mathcal E$
expressed in Eq. \eqref{eq:error_mbin1},
%
The equality is achieved
if $\dot u_i\left( q(t') \right) = c_i$
for all $i = 2, \dots, n$ at any $t'$,
up to a multiplicative factor.
%
Solving this set of equations yields
$$
\alpha(t') = \frac{              1              }
                  { \lambda_i \, (t' + t_{i0} ) },
\quad
\mathrm{with\;\;}
t_{i0} = \frac{             t            }
              { e^{ \lambda_i q(t) } - 1 }.
$$
Such a solution is possible only if
\begin{equation}
\lambda_2 = \dots = \lambda_n.
\end{equation}
%
\note{
  We add some steps for the above solution.
  Integrating $\dot u_i = c_i$ yields
  \begin{equation}
  u_i(t') = c_i \left(t' + t_{i0} \right).
  \label{eq:ui_solution}
  \end{equation}
  Given that $u_i(t) = 1$ and $u_i(0) = e^{-\lambda_i \, q(t)}$,
  we have
  $$
  c_i = \frac{ 1 }{ t + t_{i0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{i0} } { t + t_{i0} }
  =
  e^{ -\lambda_k \, q(t) }.
  $$
  Taking the logarithm of Eq. \eqref{eq:ui_solution} yields
  $$
  -\lambda_i \, q(t) + \lambda_i \, q(t')
  = \log c_i + \log\left( t' + t_{i0} \right).
  $$
  Differentiating this with respect to $t'$ yields
  $$
  \lambda_i \, \alpha(t') = \frac{ 1 } { t' + t_{i0} }.
  $$
}

@@

If we wish to find,
for a given updating matrix $\mathbf w$,
a near-optimal $\alpha(t)$,
%
then a good candidate is
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_trial}
\end{equation}
%
for some optimal value of $\lambda$.
%
Using Eq. \eqref{eq:alpha_trial}
in Eq. \eqref{eq:error_mbin1} yields
%
\begin{align}
\mathcal E
&=
\frac{1}{t+t_0}
\sum_{i=2}^n
  \frac{       \Gamma_i \, \lambda_i^2      }
       { \lambda \, (2 \lambda_i - \lambda) }
\left[
  1 - \left(
        1 + \frac{ t }{ t_0 }
      \right)^{1 - 2 \, \lambda_i/\lambda}
\right]
\notag
\\
&\stackrel{      t \gg t_0      }
          { =\joinrel=\joinrel= }
%
\sum_{i = 2}^n
\begin{dcases}
  \frac{ 1 }
       { t }
  \frac{         \Gamma_i \, \lambda_i^2       }
       { \lambda \, (2 \, \lambda_i - \lambda) }
  &
  \mathrm{if \;} \lambda < 2 \, \lambda_i,
  \\%[1em]
  %
  %
  \frac{ \Gamma_i }
       {  4 \, t  }
  \ln \frac{ t + t_0 }
           {  t_0    }
  &
  \mathrm{if \;} \lambda = 2 \, \lambda_i,
  \\%[1em]
  %
  %
  \frac{ t_0^{ 2 \, \lambda_i / \lambda  - 1 } }
       {   t^{ 2 \, \lambda_i / \lambda      } }
  \frac{      \Gamma_i \, \lambda_i^2          }
       { \lambda \, (\lambda - 2 \, \lambda_i) }
  &
  \mathrm{if \;} \lambda > 2 \, \lambda_i.
\end{dcases}
\label{eq:y2_sum_perfect_invt}
\end{align}
%
\note{In the above expression,
$t + t_0$ are replaced by $t$.
%
Adding back $t_0$, we get
$$
\begin{aligned}
  \mathcal E
  \stackrel{         t \to \infty         }
           { =\joinrel=\joinrel=\joinrel= }
  \sum_{i = 2}^n
  \begin{dcases}
    \frac{    1    }
         { t + t_0 }
    \frac{       \Gamma_i \, \lambda_i^2         }
         { \lambda \, (2 \, \lambda_i - \lambda) }
    &
    \mathrm{if \;} \lambda < 2 \, \lambda_i,
    \\%[1em]
    %
    %
    \frac{    \Gamma_i    }
         { 4 \, (t + t_0) }
    \ln \frac{ t + t_0 }
             {   t_0  }
    &
    \mathrm{if \;} \lambda = 2 \, \lambda_i,
    \\
    %
    %
    \frac{  t_0^{ 2 \, \lambda_i / \lambda  - 1}  }
         { (t + t_0)^{ 2 \, \lambda_i / \lambda } }
    \frac{      \Gamma_i \, \lambda_i^2          }
         { \lambda \, (\lambda - 2 \, \lambda_i) }
    &
    \mathrm{if \;} \lambda > 2 \, \lambda_i.
  \end{dcases}
\end{aligned}
$$
}
%
The last two cases, $\lambda \ge 2\lambda_i$,
are asymptotically suboptimal
as they decay slower than $1/t$.
%
Thus, the optimal $\lambda$
satisfies $\lambda < 2 \min \lambda_i$,
and in this case,
we can seek that
the optimal value from
the smallest real root of
%
\begin{equation}
\sum_{i = 2}^n
\frac{ \lambda_i - \lambda }
{ \left(2 - \lambda/ \lambda_i \right)^2 }
= 0.
\label{eq:optimal_lambda_approx}
\end{equation}
